Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2016-01-07 paper/paper_acm.tex
add things

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

@@

This inductive step requires to formalize the notion of left-exact
reflective subuniverse, which corresponds to a stratified version of
left-exact modality introduced in~\cite[Chapter~7]{hottbook}. A
Lawvere-Tierney topology can thus be seen as a left-exact modality on
$\Type_{-1}$ and LT-sheafification as an inductive process that extends
it as a left-exact modality on $\Type_{n}$ for any $n$.

@@

@@

Section~\ref{sec:hott} presents the foundations of homotopy type theory
and present a analogous of~\cite[Corollary 6.2.3.5]{lurie}. 
%
Section~\ref{sec:lexmod} introduces the notion of left-exact stratified
modalities, on which this paper is based.
%
Section~\ref{sec:sheaves} defines Lawvere-Tierney sheaves in homotopy
type theory.
%
Section~\ref{sec:sheafification} explains the construction of the
LT-sheafification functor.
%
The definitions of this paper uses an informal presentation of
homotopy type theory, working entirely inside of Univalent
Foundations. 
 %
Section~\ref{sec:formalization} discusses the details of the
formalization in Coq using formal type theory, in particular issues
concerning universe polymorphism.
%
Finally, Section~\ref{sec:future-works} concludes on future works.

@@

% Using $\Type_n$ instead of just $\Type$ is the first step to connect
% type theory to higher topos theory. The next step is to exhibit a
% hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}[-0.3em]
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath}%
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Again, we define some sugar. A function $f$ is 
\begin{itemize}
\item an {\em embedding} if $f$ is $(-1)$-truncated
\item a {\em surjection} is every fiber of $f$ is merely inhabited
  (\ie $\|\fib f y\|$ holds for all $y$).
\end{itemize}
Then one can show~\cite[Lemma 7.6.4]{hottbook} that any map $f$
factors uniquely through $\im(f) \defeq \sum_{y:B} \|\fib f y\|$ as a
surjection followed by an embedding.

@@

A fundamental property in the construction of the LT-sheafification
functor is the fact that the type of modal types at level $n$ defines
a sheaf.
%
This property initially holds in the topos-theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses the specific definitions
of separated types and sheaves, which has constituted a guide in the
choice of their precise statements.

@@

\begin{prop}
  $(\modal_{\lnot\lnot})_n$, the modality obtained by LT-sheafification
  of the double negation modality,
  %
  induces a type theory where the propositional excluded middle law holds.
\end{prop}

@@

@@

One issue with universe polymorphism lies in the management of
recursive definitions. Indeed, the following recursive definition of
LT-sheafification
%
\vspace{-0.05em}
\[ \begin{array}{l}
   \modal : \forall \ (n : nat), \ \Type_n \to \Type_n 
   \\
    \modal_{-1\phantom{n}}(T) \defeq\lnot\lnot T \\

      \displaystyle{\modal_{n+1}(T)} \defeq  
      \displaystyle{\sum_{u:T \to \Type_n^\modal} \!\!\!\!\modal_{-1} 
      \left\|
      \sum_{a:T} u= (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
\vspace{-0.3em}
\]
%
is not allowed. 
%
This is because Coq forces the universe of the first $\Type_n$
occurring in the definition to be the same for every $n$, whereas the
universe of the first $\Type_{n+1}$ occurring in $\modal_{n+1}$ should be at
least one level higher as the one of $\Type_n$ occurring in
$\modal_{n}$ because of the use of $\Sigma$-type over
$T \to \Type_n^\modal$ and equality on the return type of $\modal_n$.
%
% \nt{explain why it is an issue.}
%
Thus, the induction step presented in this paper has been formalized,
but the complete recursive LT-sheafification can not be defined for the
moment.
%
Note that the same increasing in the universe levels occurs in the
Rezk completion for categories~\cite{rezk}. In the definition of the
completion, they use the Yoneda embedding and representable functors,
which is similar to our use of characteristic functions.

@@

As a future work, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  The extension to whole $\Type$ in lemma~\ref{lem:type} is not
  totally satisfactory, as every type is collapsed to a truncated
  one. But some types in homotopy type theory are not
  truncated~\cite[Example 8.8.6]{hottbook}.
  % At the moment, sheafification only works with truncated
  % types. This is fine from a programming point of view, 
  % but many important types in homotopy theory are not
  % truncated, \eg the circle $\mathbb S^2$.  Thus, we plan to extend
  % the sheafification to all types by using a
  % limit construction.
\item 
  We would like to have much more examples of left-exact modalities on
  $\HProp$, in order to extend type theory with different principles
  than excluded middle.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Gro\-thendieck~\cite[Section~V.4]{maclanemoerdijk} in the sense that any
Grothen\-dieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.

>>>>

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor gro-then-dieck LT-sheafi-fi-ca-tion}

@@

Section~\ref{sec:lawv-tiern-nutshell} presents an overview of the construction. 
% 
Section~\ref{sec:hott} presents the foundations of homotopy type theory
and present a analogous of~\cite[Corollary 6.2.3.5]{lurie}. 
%
Section~\ref{sec:lexmod} introduces the notion of left-exact stratified
modalities, on which this paper is based.
%
Section~\ref{sec:sheaves} defines Lawvere-Tierney sheaves in homotopy
type theory.
%
Section~\ref{sec:sheafification} explains the construction of the
LT-sheafification functor.
%
The definitions of this paper uses an informal presentation of
homotopy type theory, working entirely inside of Univalent
Foundations. 
 %
Section~\ref{sec:formalization} discusses the details of the
formalization in Coq using formal type theory, in particular issues
concerning universe polymorphism.
%
Finally, Section~\ref{sec:future-works} concludes on future works.

\section{Lawvere-Tierney Sheafification in a nutshell}
\label{sec:lawv-tiern-nutshell}

@@

This inductive step requires to formalize the notion of left-exact
reflective subuniverse, which corresponds to a stratified version of
left-exact modality introduced in~\cite[Chapter~7]{hottbook}. A
Lawvere-Tierney topology can thus be seen as a left-exact modality on
$\Type_{-1}$ and LT-sheafification as an inductive process that extends
it as a left-exact modality on $\Type_{n}$ for any $n$.
Our inductive definition of these modalities will be given by
\[ \begin{array}{l}
   \modal : \forall \ (n : nat), \ \Type_n \to \Type_n 
   \\
    \modal_{-1\phantom{n}}(T) \defeq\lnot\lnot T \\

      \displaystyle{\modal_{n+1}(T)} \defeq  
      \displaystyle{\sum_{u:T \to \Type_n^\modal} \!\!\!\!\modal_{-1} 
      \left\|
      \sum_{a:T} u= (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
% \vspace{-0.3em}
\]
where $\Type_n^\modal$ is the type of modal types for $\modal_n$.
The rest of the paper is an inductive proof that this definition produces a
sequence of left-exact modalities.

@@

Such a construction can be thought in two ways:
\begin{itemize}
\item as a translation from one type theory into another type theory
  satisfying new principles
\item or as a transformation, taking a model of type theory, and
  producing another model, where the desired principle is true. This
  can lead to consistency results.
\end{itemize}

@@

@@

% Using $\Type_n$ instead of just $\Type$ is the first step to connect
% type theory to higher topos theory. The next step is to exhibit a
% hierarchy of subobject classifiers on $n$-truncated homotopy fibers.
%
The homotopy fiber $\fib{f}{b}$ of a function $f$ at element $b$ is
defined as 
\begin{mymath}[-0.3em]
\fib{f}{b} \defeq \sum_{a:A} f(a) = b.
\end{mymath}%
%
A function $f$ is with $n$-truncated homotopy fibers (or simply
$n$-truncated function) when $\fib{f}{b}$
is in $\Type_n$ for any $b$.  
%
Again, we define some sugar. A function $f$ is 
\begin{itemize}
\item an {\em embedding} if $f$ is $(-1)$-truncated
\item a {\em surjection} if every fiber of $f$ is merely inhabited
  (i.e $\|\fib f y\|$ holds for all $y$).
\end{itemize}
Then one can show~\cite[Lemma 7.6.4]{hottbook} that any map $f$
factors uniquely through $\im(f) \defeq \sum_{y:B} \|\fib f y\|$ as a
surjection followed by an embedding.

@@

A fundamental property in the construction of the LT-sheafifica\-tion
functor is the fact that the type of modal types at level $n$ defines
a sheaf.
%
This property initially holds in the topos-theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses the specific definitions
of separated types and sheaves, which has constituted a guide in the
choice of their precise statements.

@@

\begin{prop}
  $(\modal_{\lnot\lnot})_n$, the modality obtained by LT-sheafifica\-tion
  of the double negation modality,
  %
  induces a type theory where the propositional excluded middle law holds.
\end{prop}

@@

We can view sheafification in terms of model of type theory ; let
$\mathfrak M$ be a model of homotopy type theory. Using sheafification
associated to a truncation index $n$, one can build a modality
$\modal$, which is left-exact with respect to types at most
$n$-truncated. Thus, there is a model $\modal \mathfrak M$ of type
theory (using results in section~\ref{sec:new-type-theory}), where
excluded middle is true, and which is univalent when restricted to $n$-types.

@@

One issue with universe polymorphism lies in the management of
recursive definitions. Indeed, the following recursive definition of
LT-sheafification
%
% \vspace{-0.05em}
\[ \begin{array}{l}
   \modal : \forall \ (n : nat), \ \Type_n \to \Type_n 
   \\
    \modal_{-1\phantom{n}}(T) \defeq\lnot\lnot T \\

      \displaystyle{\modal_{n+1}(T)} \defeq  
      \displaystyle{\sum_{u:T \to \Type_n^\modal} \!\!\!\!\modal_{-1} 
      \left\|
      \sum_{a:T} u= (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
% \vspace{-0.3em}
\]
%
is not allowed. 
%
This is because Coq forces the universe of the first $\Type_n$
occurring in the definition to be the same for every $n$, whereas the
universe of the first $\Type_{n+1}$ occurring in $\modal_{n+1}$ should be at
least one level higher as the one of $\Type_n$ occurring in
$\modal_{n}$ because of the use of $\Sigma$-type over
$T \to \Type_n^\modal$ and equality on the return type of $\modal_n$.
%
% \nt{explain why it is an issue.}
%
Thus, the induction step presented in this paper has been formalized,
but the complete recursive LT-sheafification can not be defined for the
moment.
%
Note that the same increasing in the universe levels occurs in the
Rezk completion for categories~\cite{rezk}. In the definition of the
completion, they use the Yoneda embedding and representable functors,
which is similar to our use of characteristic functions.

@@

As a future work, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  The extension to whole $\Type$ in lemma~\ref{lem:type} is not
  totally satisfactory, as every type is collapsed to a truncated
  one. But some types in homotopy type theory are not
  truncated~\cite[Example 8.8.6]{hottbook}.
  % At the moment, sheafification only works with truncated
  % types. This is fine from a programming point of view, 
  % but many important types in homotopy theory are not
  % truncated, \eg the circle $\mathbb S^2$.  Thus, we plan to extend
  % the sheafification to all types by using a
  % limit construction.
\item 
  We would like to have much more examples of left-exact modalities on
  $\HProp$, in order to extend type theory with different principles
  than excluded middle.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Grothen\-dieck~\cite[Section~V.4]{maclanemoerdijk} in the sense that any
Grothen\-dieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.
